---
title: "sp_f_combined"
date: "2026-03-12"
author: "Celeste Basken"
output: html_document
---

This is the combined SP26 + F25 data in a new optimization.

Now to go to Crazy Town and use SP26 data too!!

# Part 0: Load data

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(lpSolve)
library(readr)
library(tidyr)

optimization2 <- read.csv("f25sp26_unknowns_removed.csv", stringsAsFactors = FALSE)

optimization2 <- optimization2 %>%
  mutate(across(
    c(Sustainable_price_lb,
      Conventional_price_lb,
      Expected_lb_meat_portion,
      Conventional_GHG_per_lb,
      Baseline_freq),
    ~replace_na(parse_number(as.character(.)), 0)
  ))

```

# Part 1

Fall + Spring 26 menus ASSUMING that we

* Delete any meals with fewer than 100 portions forecasted
* Replace ingredients that I can’t find with what I assume to be appropriate substitutes?

Just a note: Columns with dif names than 2v2 (optimization csv)

* Menu_freq_f25 -> Baseline_freq
* %_dish_meat -> Percent_dish_meat

## Scenario s1v2.2: fall n spring with unknowns removed 3.10.26

```{r}
# =========================
# s1v2.2: Max cost reduction WITHOUT losing sustainability
# (Primary: keep sustainable spend >= baseline)
# (Secondary: minimize total cost)
# Optional tertiary: minimize GHG (after finding min cost)
# =========================

# Setup
# ----------------------------
x0 <- optimization2$Baseline_freq
n  <- nrow(optimization2)

meals_served <- sum(x0)

# Lower bound (e.g., max 50% reduction)
reduction <- 0.25
lower <- ceiling(x0 * reduction)

# Upper bound (e.g., max 50% increase)
max_multiplier <- 1.25
upper <- floor(x0 * max_multiplier)

# ----------------------------
# NEW: portion-adjusted price per "menu appearance"
# ----------------------------
# Clean/validate portion weights
if (!("Expected_lb_meat_portion" %in% names(optimization2))) {
  stop("Column Expected_lb_meat_portion not found in optimization2_data.csv")
}

portion_lb <- optimization2$Expected_lb_meat_portion
if (any(is.na(portion_lb))) {
  stop("Expected_lb_meat_portion has NA values. Please fill them before optimizing.")
}
if (any(portion_lb < 0)) {
  stop("Expected_lb_meat_portion has negative values. Please correct them.")
}

# ---- portion-adjusted vectors ----
portion_lb <- optimization2$Expected_lb_meat_portion
default_sus_clean <- tolower(trimws(as.character(optimization2$Default_sus)))
sus_flag <- as.integer(default_sus_clean == "yes")

# Effective price per lb used for costing:
# if Default_sus == yes -> Sustainable_price_lb else Conventional_price_lb
price_lb <- ifelse(sus_flag == 1,
                   optimization2$Sustainable_price_lb,
                   optimization2$Conventional_price_lb)

# Portion-adjusted "cost per appearance" (THIS is what we minimize / constrain on)
cost_per_appearance <- price_lb * portion_lb

# "Sustainable spend per appearance" (only counts when Default_sus == yes)
sus_cost_per_appearance <- optimization2$Sustainable_price_lb * portion_lb * sus_flag

# GHG per appearance (uses your existing conventional GHG per lb * portion lbs)
ghg_per_appearance <- optimization2$Conventional_GHG_per_lb * portion_lb

# ----------------------------
# Constraint matrix
# ----------------------------
# 1) Maintain total meals: sum(x) = meals_served
A <- matrix(1, nrow = 1, ncol = n)
dir <- "="
rhs <- meals_served

# 2) Lower bounds: x_i >= lower_i
A <- rbind(A, diag(n))
dir <- c(dir, rep(">=", n))
rhs <- c(rhs, lower)

# 3) Upper bounds: x_i <= upper_i (for non-NA uppers)
if (any(!is.na(upper))) {
  idx <- which(!is.na(upper))
  A <- rbind(A, diag(n)[idx, , drop = FALSE])
  dir <- c(dir, rep("<=", length(idx)))
  rhs <- c(rhs, upper[idx])
}

# ---- baseline totals under new accounting ----
baseline_cost <- sum(cost_per_appearance * x0)
baseline_sus  <- sum(sus_cost_per_appearance * x0)

# Tuning knob: allow tiny wiggle if you want (0 = strict "no loss")
sus_slack <- 0.0   # e.g., 0.001 would allow 0.1% loss

# ============================================================
# Add sustainability floor constraint (NO sustainability loss)
# ============================================================
A_susfloor   <- rbind(A, sus_cost_per_appearance)
dir_susfloor <- c(dir, ">=")
rhs_susfloor <- c(rhs, baseline_sus * (1 - sus_slack))

# ============================================================
# Stage 1: Minimize cost subject to sustainability floor
# (This gives "maximum cost reduction without losing sustainability")
# ============================================================
sol_s1v2.2_stage1 <- lp(
  direction = "min",
  objective.in = cost_per_appearance,
  const.mat = A_susfloor,
  const.dir = dir_susfloor,
  const.rhs = rhs_susfloor,
  all.int = TRUE
)

if (sol_s1v2.2_stage1$status != 0) {
  stop("S1v2.1 Stage 1 infeasible. Try relaxing bounds or set sus_slack to a small value (e.g., 0.001).")
}

x_s1v2.2_cost <- sol_s1v2.2_stage1$solution
opt_cost_s1v2.2 <- sum(cost_per_appearance * x_s1v2.2_cost)

# ============================================================
# Optional Stage 2: Minimize GHG while keeping:
#  - cost at its minimum (within tiny slack)
#  - sustainability >= baseline
# ============================================================
cost_slack <- 0.001  # allow tiny rounding wiggle around min cost

A_costcap2   <- rbind(A_susfloor, cost_per_appearance)
dir_costcap2 <- c(dir_susfloor, "<=")
rhs_costcap2 <- c(rhs_susfloor, opt_cost_s1v2.2 * (1 + cost_slack))

sol_s1v2.2_stage2 <- lp(
  direction = "min",
  objective.in = ghg_per_appearance,
  const.mat = A_costcap2,
  const.dir = dir_costcap2,
  const.rhs = rhs_costcap2,
  all.int = TRUE
)

if (sol_s1v2.2_stage2$status != 0) {
  stop("S1v2.1 Stage 2 infeasible. Try increasing cost_slack slightly (e.g., 0.005).")
}

x_s1v2.2_final <- sol_s1v2.2_stage2$solution

# ============================================================
# Summaries
# ============================================================
summ <- function(x) {
  total_cost <- sum(cost_per_appearance * x)
  sus_spend  <- sum(sus_cost_per_appearance * x)
  sus_pct    <- if (total_cost > 0) sus_spend / total_cost else NA_real_
  total_ghg  <- sum(ghg_per_appearance * x)
  c(meals = sum(x), total_cost = total_cost, sus_spend = sus_spend, sus_pct = sus_pct, total_ghg = total_ghg)
}

s1v2.2_table <- rbind(
  baseline = summ(x0),
  s1v2_1_stage1_min_cost_no_sus_loss = summ(x_s1v2.2_cost),
  s1v2_1_final_min_ghg_given_cost_and_no_sus_loss = summ(x_s1v2.2_final)
)

print(s1v2.2_table)

# Attach frequencies
optimization2$freq_opt_s1v2.2 <- x_s1v2.2_final
optimization2$delta_s1v2.2 <- optimization2$freq_opt_s1v2.2 - optimization2$Baseline_freq
```



```{r}

# -----------------------------------------
# Add optimized frequencies to dataframe
# -----------------------------------------

optimization2 <- optimization2 %>%
  mutate(
    freq_baseline = Baseline_freq,
    freq_opt_s1v2_2 = x_s1v2.2_final,
    delta_s1v2_2 = freq_opt_s1v2_2 - freq_baseline
  )


# -----------------------------------------
# Category totals
# -----------------------------------------

category_totals_s1v2_2 <- optimization2 %>%
  group_by(Category) %>%
  summarise(
    baseline_meals = sum(freq_baseline, na.rm = TRUE),
    opt_meals_s1v2_2 = sum(freq_opt_s1v2_2, na.rm = TRUE),
    delta_meals_s1v2_2 = sum(delta_s1v2_2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    baseline_share = baseline_meals / sum(baseline_meals, na.rm = TRUE),
    opt_share_s1v2_2 = opt_meals_s1v2_2 / sum(opt_meals_s1v2_2, na.rm = TRUE)
  ) %>%
  arrange(desc(opt_meals_s1v2_2))


# View category totals
category_totals_s1v2_2

```


```{r}
# -----------------------------------------
# Save updated optimization2 table
# -----------------------------------------

write.csv(
  optimization2,
  file = "sp_f_combined_outputs/optimization_s1v2_2.csv",
  row.names = FALSE
)
```

## Scenario 2v2.2 
Ugh I hate this naming convention but it's my own fault

```{r}
# =========================
# s2v2.2: Max sustainability WITHOUT increasing cost
# (Primary: maximize sustainable spend)
# (Secondary constraint: total cost <= baseline)
# Optional tertiary: minimize GHG (after finding max sustainability)
# =========================

# Setup
# ----------------------------
x0 <- optimization2$Baseline_freq
n  <- nrow(optimization2)

meals_served <- sum(x0)

# Lower bound (e.g., max 75% reduction if reduction = 0.25)
reduction <- 0.25
lower <- ceiling(x0 * reduction)

# Upper bound (e.g., max 25% increase)
max_multiplier <- 1.25
upper <- floor(x0 * max_multiplier)

# ----------------------------
# Portion-adjusted price per "menu appearance"
# ----------------------------
if (!("Expected_lb_meat_portion" %in% names(optimization2))) {
  stop("Column Expected_lb_meat_portion not found in optimization2")
}

portion_lb <- optimization2$Expected_lb_meat_portion
if (any(is.na(portion_lb))) {
  stop("Expected_lb_meat_portion has NA values. Please fill them before optimizing.")
}
if (any(portion_lb < 0)) {
  stop("Expected_lb_meat_portion has negative values. Please correct them.")
}

default_sus_clean <- tolower(trimws(as.character(optimization2$Default_sus)))
sus_flag <- as.integer(default_sus_clean == "yes")

# Effective price per lb used for costing:
# if Default_sus == yes -> Sustainable_price_lb else Conventional_price_lb
price_lb <- ifelse(sus_flag == 1,
                   optimization2$Sustainable_price_lb,
                   optimization2$Conventional_price_lb)

# Portion-adjusted vectors
cost_per_appearance <- price_lb * portion_lb

# Sustainable spend per appearance
sus_cost_per_appearance <- optimization2$Sustainable_price_lb * portion_lb * sus_flag

# GHG per appearance
ghg_per_appearance <- optimization2$Conventional_GHG_per_lb * portion_lb

# ----------------------------
# Constraint matrix
# ----------------------------
# 1) Maintain total meals: sum(x) = meals_served
A <- matrix(1, nrow = 1, ncol = n)
dir <- "="
rhs <- meals_served

# 2) Lower bounds: x_i >= lower_i
A <- rbind(A, diag(n))
dir <- c(dir, rep(">=", n))
rhs <- c(rhs, lower)

# 3) Upper bounds: x_i <= upper_i
if (any(!is.na(upper))) {
  idx <- which(!is.na(upper))
  A <- rbind(A, diag(n)[idx, , drop = FALSE])
  dir <- c(dir, rep("<=", length(idx)))
  rhs <- c(rhs, upper[idx])
}

# ----------------------------
# Baseline totals
# ----------------------------
baseline_cost <- sum(cost_per_appearance * x0)
baseline_sus  <- sum(sus_cost_per_appearance * x0)

# Tuning knob: allow tiny cost wiggle if desired
cost_slack <- 0.0   # strict no increase; e.g. 0.001 allows 0.1% increase

# ============================================================
# Add cost cap constraint (NO cost increase)
# ============================================================
A_costcap   <- rbind(A, cost_per_appearance)
dir_costcap <- c(dir, "<=")
rhs_costcap <- c(rhs, baseline_cost * (1 + cost_slack))

# ============================================================
# Stage 1: Maximize sustainability subject to cost cap
# ============================================================
sol_s2v2.2_stage1 <- lp(
  direction = "max",
  objective.in = sus_cost_per_appearance,
  const.mat = A_costcap,
  const.dir = dir_costcap,
  const.rhs = rhs_costcap,
  all.int = TRUE
)

if (sol_s2v2.2_stage1$status != 0) {
  stop("S2v2.2 Stage 1 infeasible. Try relaxing bounds or increasing cost_slack slightly.")
}

x_s2v2.2_sus <- sol_s2v2.2_stage1$solution
opt_sus_s2v2.2 <- sum(sus_cost_per_appearance * x_s2v2.2_sus)

# ============================================================
# Optional Stage 2: Minimize GHG while keeping:
#  - sustainability at its maximum (within tiny slack)
#  - cost <= baseline
# ============================================================
sus_slack <- 0.001  # allow tiny rounding wiggle around max sustainability

A_ghg2   <- rbind(A_costcap, sus_cost_per_appearance)
dir_ghg2 <- c(dir_costcap, ">=")
rhs_ghg2 <- c(rhs_costcap, opt_sus_s2v2.2 * (1 - sus_slack))

sol_s2v2.2_stage2 <- lp(
  direction = "min",
  objective.in = ghg_per_appearance,
  const.mat = A_ghg2,
  const.dir = dir_ghg2,
  const.rhs = rhs_ghg2,
  all.int = TRUE
)

if (sol_s2v2.2_stage2$status != 0) {
  stop("S2v2.2 Stage 2 infeasible. Try increasing sus_slack slightly (e.g., 0.005).")
}

x_s2v2.2_final <- sol_s2v2.2_stage2$solution

# ============================================================
# Summaries
# ============================================================
summ <- function(x) {
  total_cost <- sum(cost_per_appearance * x)
  sus_spend  <- sum(sus_cost_per_appearance * x)
  sus_pct    <- if (total_cost > 0) sus_spend / total_cost else NA_real_
  total_ghg  <- sum(ghg_per_appearance * x)
  c(meals = sum(x), total_cost = total_cost, sus_spend = sus_spend, sus_pct = sus_pct, total_ghg = total_ghg)
}

s2v2.2_table <- rbind(
  baseline = summ(x0),
  s2v2_2_stage1_max_sus_no_cost_increase = summ(x_s2v2.2_sus),
  s2v2_2_final_min_ghg_given_max_sus_and_no_cost_increase = summ(x_s2v2.2_final)
)

print(s2v2.2_table)

# Attach frequencies
optimization2$freq_opt_s2v2.2 <- x_s2v2.2_final
optimization2$delta_s2v2.2 <- optimization2$freq_opt_s2v2.2 - optimization2$Baseline_freq
```




```{r}
# -----------------------------------------
# Add optimized frequencies to dataframe
# -----------------------------------------

optimization2 <- optimization2 %>%
  mutate(
    freq_baseline = Baseline_freq,
    freq_opt_s2v2_2 = x_s2v2.2_final,
    delta_s2v2_2 = freq_opt_s2v2_2 - freq_baseline
  )


# -----------------------------------------
# Category totals
# -----------------------------------------

category_totals_s2v2_2 <- optimization2 %>%
  group_by(Category) %>%
  summarise(
    baseline_meals = sum(freq_baseline, na.rm = TRUE),
    opt_meals_s2v2_2 = sum(freq_opt_s2v2_2, na.rm = TRUE),
    delta_meals_s2v2_2 = sum(delta_s2v2_2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    baseline_share = baseline_meals / sum(baseline_meals, na.rm = TRUE),
    opt_share_s2v2_2 = opt_meals_s2v2_2 / sum(opt_meals_s2v2_2, na.rm = TRUE)
  ) %>%
  arrange(desc(opt_meals_s2v2_2))


# View category totals
category_totals_s2v2_2

```


```{r}
# -----------------------------------------
# Save updated optimization2 table
# -----------------------------------------

write.csv(
  optimization2,
  file = "sp_f_combined_outputs/optimization_s2v2_2.csv",
  row.names = FALSE
)
```

