---
title: "SP_F_C3"
output: html_document
---

We're heading back to Crazy Town and doing this like I always should have, aka by generating the complete report endogenously

# Part 0: Load/ Build data

## Packages

```{r}
library(dplyr)
library(readr)
library(janitor)
library(lpSolve)

```


## Read Data

```{r}
safe_mean <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

first_nonmissing <- function(x) {
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) NA_character_ else x[1]
}

# -------------------------
# Read + clean
# -------------------------
meals <- read.csv("../Basic_Data/F25_Sp26_meals.csv", stringsAsFactors = FALSE) %>%
  clean_names()

ingredient_prices <- read.csv("../Basic_Data/Ingredient_prices.csv", stringsAsFactors = FALSE) %>%
  clean_names()

ghg_equivalents <- read.csv("../Basic_Data/GHG_equivalents.csv", stringsAsFactors = FALSE) %>%
  clean_names()

# -------------------------
# Standardize values
# -------------------------
meals <- meals %>%
  mutate(
    dining_hall = trimws(dining_hall),
    ingredient = trimws(ingredient),
    category = trimws(category),
    planned_portions = as.numeric(gsub(",", "", planned_portions)),
    percent_meat = as.numeric(percent_meat),
    meat_price_per_dish = as.numeric(meat_price_per_dish),
    expected_lb_meat_portion = as.numeric(expected_lb_meat_portion),
    oz_meat_per_dish = as.numeric(oz_meat_per_dish),
    portion_size_oz = as.numeric(portion_size_oz),
    planned_weight_lbs = as.numeric(planned_weight_lbs),
    cost_recipe_per_portion = as.numeric(cost_recipe_per_portion)
  )

ingredient_prices <- ingredient_prices %>%
  mutate(
    ingredient = trimws(ingredient),
    category = trimws(category),
    conventional_price_lb = as.numeric(conventional_price_lb),
    sustainable_price_lb = as.numeric(sustainable_price_lb),
    default_sus = trimws(default_sus)
  )

ghg_equivalents <- ghg_equivalents %>%
  mutate(
    meat_type = trimws(meat_type),
    c_footprint_kg_c_per_kg_food = as.numeric(c_footprint_kg_c_per_kg_food)
  )

```

## Building optimization df

This is what I used to just read in as ready for r

```{r}
# =========================================================
# Build optimization_data_cafe3
# =========================================================

meals_cafe3 <- meals %>%
  filter(dining_hall == "C3")

ingredient_summary_cafe3 <- meals_cafe3 %>%
  filter(!is.na(ingredient), ingredient != "") %>%
  group_by(ingredient) %>%
  summarise(
    baseline_freq = n(),
    percent_dish_meat = safe_mean(percent_meat),
    meat_price_per_dish = safe_mean(meat_price_per_dish),
    expected_lb_meat_portion = safe_mean(expected_lb_meat_portion),
    oz_meat_per_dish = safe_mean(oz_meat_per_dish),
    portion_size_oz = safe_mean(portion_size_oz),
    expected_portions = safe_mean(planned_portions),
    planned_weight_lbs = safe_mean(planned_weight_lbs),
    cost_recipe_per_portion = safe_mean(cost_recipe_per_portion),
    category_meals = first_nonmissing(category),
    .groups = "drop"
  )

ingredient_lookup <- ingredient_prices %>%
  distinct(ingredient, .keep_all = TRUE) %>%
  transmute(
    ingredient,
    category_prices = category,
    conventional_price_lb,
    sustainable_price_lb,
    default_sus
  )

ghg_lookup <- ghg_equivalents %>%
  distinct(meat_type, .keep_all = TRUE) %>%
  transmute(
    category = meat_type,
    conventional_ghg_per_lb = c_footprint_kg_c_per_kg_food
  )

optimization_data_cafe3 <- ingredient_summary_cafe3 %>%
  left_join(ingredient_lookup, by = "ingredient") %>%
  mutate(
    category = coalesce(category_prices, category_meals)
  ) %>%
  left_join(ghg_lookup, by = "category") %>%
  select(
    ingredient,
    category,
    percent_dish_meat,
    meat_price_per_dish,
    expected_lb_meat_portion,
    oz_meat_per_dish,
    conventional_price_lb,
    sustainable_price_lb,
    default_sus,
    portion_size_oz,
    expected_portions,
    planned_weight_lbs,
    cost_recipe_per_portion,
    conventional_ghg_per_lb,
    baseline_freq
  )

optimization_data_cafe3 <- optimization_data_cafe3 %>%
  mutate(
    baseline_freq = as.numeric(baseline_freq)
  )

# -------------------------
# Quick diagnostics
# -------------------------
optimization_data_cafe3 %>%
  filter(
    is.na(expected_lb_meat_portion) |
    is.na(conventional_price_lb) |
    is.na(sustainable_price_lb) |
    is.na(default_sus) |
    is.na(conventional_ghg_per_lb)
  )

```

# Scenario s1_cafe3

Maximize cost cutting (minimize cost) while holding sustainability

```{r}
# =========================================================
# s1_cafe3: Max cost reduction WITHOUT losing sustainability
# Primary: keep sustainable spend >= baseline
# Secondary: minimize total cost
# Optional tertiary summary outputs included
# =========================================================

opt <- optimization_data_cafe3

# -------------------------
# Setup
# -------------------------
x0 <- opt$baseline_freq
n  <- nrow(opt)

meals_served <- sum(x0)

# Lower bound (max 25% reduction per ingredient)
reduction <- 0.5
lower <- ceiling(x0 * reduction)

# Upper bound (max 25% increase per ingredient)
max_multiplier <- 1.5
upper <- floor(x0 * max_multiplier)

# -------------------------
# Feasibility checks
# -------------------------
required_cols <- c(
  "expected_lb_meat_portion",
  "conventional_price_lb",
  "sustainable_price_lb",
  "default_sus",
  "conventional_ghg_per_lb",
  "baseline_freq"
)

missing_cols <- setdiff(required_cols, names(opt))
if (length(missing_cols) > 0) {
  stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))
}

if (any(is.na(opt$expected_lb_meat_portion))) {
  stop("expected_lb_meat_portion has NA values in optimization_data_cafe3.")
}
if (any(opt$expected_lb_meat_portion < 0)) {
  stop("expected_lb_meat_portion has negative values in optimization_data_cafe3.")
}
if (any(is.na(opt$default_sus))) {
  stop("default_sus has NA values in optimization_data_cafe3.")
}
if (any(is.na(opt$conventional_ghg_per_lb))) {
  stop("conventional_ghg_per_lb has NA values in optimization_data_cafe3.")
}

default_sus_clean <- tolower(trimws(as.character(opt$default_sus)))

if (any(!(default_sus_clean %in% c("yes", "no")))) {
  stop("default_sus contains values other than Yes/No.")
}

sus_flag <- as.integer(default_sus_clean == "yes")
portion_lb <- opt$expected_lb_meat_portion

if (any(sus_flag == 1 & is.na(opt$sustainable_price_lb))) {
  stop("Some default-sustainable ingredients are missing sustainable_price_lb.")
}
if (any(sus_flag == 0 & is.na(opt$conventional_price_lb))) {
  stop("Some default-conventional ingredients are missing conventional_price_lb.")
}

# -------------------------
# Per-appearance values
# -------------------------
price_lb <- dplyr::case_when(
  sus_flag == 1 ~ opt$sustainable_price_lb,
  sus_flag == 0 ~ opt$conventional_price_lb,
  TRUE ~ NA_real_
)

cost_per_appearance <- price_lb * portion_lb

sus_cost_per_appearance <- dplyr::case_when(
  sus_flag == 1 ~ opt$sustainable_price_lb * portion_lb,
  sus_flag == 0 ~ 0,
  TRUE ~ NA_real_
)

ghg_per_appearance <- opt$conventional_ghg_per_lb * portion_lb

# -------------------------
# Final safety checks
# -------------------------
if (any(is.na(x0) | is.nan(x0) | is.infinite(x0))) {
  stop("baseline_freq contains NA/NaN/Inf.")
}
if (any(is.na(lower) | is.nan(lower) | is.infinite(lower))) {
  stop("lower contains NA/NaN/Inf.")
}
if (any(is.na(upper) | is.nan(upper) | is.infinite(upper))) {
  stop("upper contains NA/NaN/Inf.")
}
if (any(is.na(cost_per_appearance) | is.nan(cost_per_appearance) | is.infinite(cost_per_appearance))) {
  stop("cost_per_appearance contains NA/NaN/Inf.")
}
if (any(is.na(sus_cost_per_appearance) | is.nan(sus_cost_per_appearance) | is.infinite(sus_cost_per_appearance))) {
  stop("sus_cost_per_appearance contains NA/NaN/Inf.")
}
if (any(is.na(ghg_per_appearance) | is.nan(ghg_per_appearance) | is.infinite(ghg_per_appearance))) {
  stop("ghg_per_appearance contains NA/NaN/Inf.")
}

# -------------------------
# Baseline totals
# -------------------------
baseline_cost <- sum(cost_per_appearance * x0)
baseline_sus  <- sum(sus_cost_per_appearance * x0)
baseline_ghg  <- sum(ghg_per_appearance * x0)

# -------------------------
# Constraint matrix
# 1) Maintain total meals
# 2) Lower bounds
# 3) Upper bounds
# 4) Sustainable spend >= baseline
# -------------------------
A <- matrix(1, nrow = 1, ncol = n)
dir <- "="
rhs <- meals_served

A <- rbind(A, diag(n))
dir <- c(dir, rep(">=", n))
rhs <- c(rhs, lower)

if (any(!is.na(upper))) {
  idx <- which(!is.na(upper))
  A <- rbind(A, diag(n)[idx, , drop = FALSE])
  dir <- c(dir, rep("<=", length(idx)))
  rhs <- c(rhs, upper[idx])
}

# Add sustainability floor
A_sus   <- rbind(A, sus_cost_per_appearance)
dir_sus <- c(dir, ">=")
rhs_sus <- c(rhs, baseline_sus)

# Coerce safely for lpSolve
A_sus_num <- matrix(
  as.numeric(A_sus),
  nrow = nrow(A_sus),
  ncol = ncol(A_sus)
)

rhs_sus_num <- as.numeric(rhs_sus)
dir_sus_chr <- as.character(dir_sus)
objective_cost <- as.numeric(cost_per_appearance)

if (any(is.na(A_sus_num) | is.nan(A_sus_num) | is.infinite(A_sus_num))) {
  stop("A_sus_num contains NA/NaN/Inf.")
}
if (any(is.na(rhs_sus_num) | is.nan(rhs_sus_num) | is.infinite(rhs_sus_num))) {
  stop("rhs_sus_num contains NA/NaN/Inf.")
}
if (any(is.na(objective_cost) | is.nan(objective_cost) | is.infinite(objective_cost))) {
  stop("objective_cost contains NA/NaN/Inf.")
}

# =========================================================
# Stage 1: Minimize total cost while keeping sustainable spend
# at or above baseline
# =========================================================
sol_s1_cafe3 <- lp(
  direction = "min",
  objective.in = objective_cost,
  const.mat = A_sus_num,
  const.dir = dir_sus_chr,
  const.rhs = rhs_sus_num,
  all.int = TRUE
)

if (sol_s1_cafe3$status != 0) {
  stop("s1_cafe3 infeasible. Try relaxing bounds or checking baseline sustainable spend.")
}

x_s1_cafe3 <- sol_s1_cafe3$solution

# =========================================================
# Summaries
# =========================================================
summ <- function(x) {
  total_cost <- sum(cost_per_appearance * x)
  sus_spend  <- sum(sus_cost_per_appearance * x)
  sus_pct    <- if (total_cost > 0) sus_spend / total_cost else NA_real_
  total_ghg  <- sum(ghg_per_appearance * x)

  c(
    meals = sum(x),
    total_cost = total_cost,
    sus_spend = sus_spend,
    sus_pct = sus_pct,
    total_ghg = total_ghg
  )
}

s1_cafe3_table <- rbind(
  baseline = summ(x0),
  s1_cafe3 = summ(x_s1_cafe3)
)

print(s1_cafe3_table)

# -------------------------
# Helpful headline metrics
# -------------------------
opt_cost_s1_cafe3 <- sum(cost_per_appearance * x_s1_cafe3)
opt_sus_s1_cafe3  <- sum(sus_cost_per_appearance * x_s1_cafe3)
opt_ghg_s1_cafe3  <- sum(ghg_per_appearance * x_s1_cafe3)

pct_cost_reduction_s1_cafe3 <- if (baseline_cost > 0) {
  100 * (baseline_cost - opt_cost_s1_cafe3) / baseline_cost
} else {
  NA_real_
}

cat("\nBaseline cost: ", baseline_cost, "\n", sep = "")
cat("Optimized cost (s1_cafe3): ", opt_cost_s1_cafe3, "\n", sep = "")
cat("Percent cost reduction achieved: ", round(pct_cost_reduction_s1_cafe3, 2), "%\n", sep = "")
cat("Baseline sustainable spend: ", baseline_sus, "\n", sep = "")
cat("Optimized sustainable spend: ", opt_sus_s1_cafe3, "\n", sep = "")
cat("Baseline GHG: ", baseline_ghg, "\n", sep = "")
cat("Optimized GHG: ", opt_ghg_s1_cafe3, "\n", sep = "")

# =========================================================
# Attach optimization outputs
# =========================================================
optimization_data_cafe3 <- optimization_data_cafe3 %>%
  mutate(
    freq_baseline = baseline_freq,
    freq_opt_s1_cafe3 = x_s1_cafe3,
    delta_s1_cafe3 = freq_opt_s1_cafe3 - freq_baseline,

    baseline_cost_contribution = freq_baseline * cost_per_appearance,
    opt_cost_contribution_s1_cafe3 = freq_opt_s1_cafe3 * cost_per_appearance,

    baseline_sus_spend_contribution = freq_baseline * sus_cost_per_appearance,
    opt_sus_spend_contribution_s1_cafe3 = freq_opt_s1_cafe3 * sus_cost_per_appearance,

    baseline_ghg_contribution = freq_baseline * ghg_per_appearance,
    opt_ghg_contribution_s1_cafe3 = freq_opt_s1_cafe3 * ghg_per_appearance
  )

# =========================================================
# Category totals for s1_cafe3
# =========================================================
category_totals_s1_cafe3 <- optimization_data_cafe3 %>%
  group_by(category) %>%
  summarise(
    baseline_meals = sum(freq_baseline, na.rm = TRUE),
    opt_meals_s1_cafe3 = sum(freq_opt_s1_cafe3, na.rm = TRUE),
    delta_meals_s1_cafe3 = opt_meals_s1_cafe3 - baseline_meals,
    pct_change_meals_s1_cafe3 = ifelse(
      baseline_meals > 0,
      (delta_meals_s1_cafe3 / baseline_meals) * 100,
      NA_real_
    ),

    baseline_share = baseline_meals / sum(optimization_data_cafe3$freq_baseline, na.rm = TRUE),
    opt_share_s1_cafe3 = opt_meals_s1_cafe3 / sum(optimization_data_cafe3$freq_opt_s1_cafe3, na.rm = TRUE),

    baseline_cost = sum(baseline_cost_contribution, na.rm = TRUE),
    opt_cost_s1_cafe3 = sum(opt_cost_contribution_s1_cafe3, na.rm = TRUE),
    delta_cost_s1_cafe3 = opt_cost_s1_cafe3 - baseline_cost,
    pct_change_cost_s1_cafe3 = ifelse(
      baseline_cost > 0,
      (delta_cost_s1_cafe3 / baseline_cost) * 100,
      NA_real_
    ),

    baseline_sus_spend = sum(baseline_sus_spend_contribution, na.rm = TRUE),
    opt_sus_spend_s1_cafe3 = sum(opt_sus_spend_contribution_s1_cafe3, na.rm = TRUE),
    delta_sus_spend_s1_cafe3 = opt_sus_spend_s1_cafe3 - baseline_sus_spend,
    pct_change_sus_spend_s1_cafe3 = ifelse(
      baseline_sus_spend > 0,
      (delta_sus_spend_s1_cafe3 / baseline_sus_spend) * 100,
      NA_real_
    ),

    baseline_ghg = sum(baseline_ghg_contribution, na.rm = TRUE),
    opt_ghg_s1_cafe3 = sum(opt_ghg_contribution_s1_cafe3, na.rm = TRUE),
    delta_ghg_s1_cafe3 = opt_ghg_s1_cafe3 - baseline_ghg,
    pct_change_ghg_s1_cafe3 = ifelse(
      baseline_ghg > 0,
      (delta_ghg_s1_cafe3 / baseline_ghg) * 100,
      NA_real_
    ),

    baseline_sus_pct = ifelse(
      baseline_cost > 0,
      baseline_sus_spend / baseline_cost,
      NA_real_
    ),
    opt_sus_pct_s1_cafe3 = ifelse(
      opt_cost_s1_cafe3 > 0,
      opt_sus_spend_s1_cafe3 / opt_cost_s1_cafe3,
      NA_real_
    ),
    delta_sus_pct_s1_cafe3 = opt_sus_pct_s1_cafe3 - baseline_sus_pct,

    .groups = "drop"
  ) %>%
  arrange(desc(opt_meals_s1_cafe3))

category_totals_s1_cafe3

```



# Scenario s5_cafe3

This is 10% cost reduction 

```{r}
# =========================================================
# s5_cafe3: Cut cost 10%, then maximize sustainability,
# then minimize GHG
# =========================================================

opt <- optimization_data_cafe3

x0 <- opt$baseline_freq
n  <- nrow(opt)

meals_served <- sum(x0)

# Lower and upper bounds
reduction <- 0.5
lower <- ceiling(x0 * reduction)

max_multiplier <- 1.5
upper <- floor(x0 * max_multiplier)

# -------------------------
# Feasibility checks
# -------------------------
required_cols <- c(
  "expected_lb_meat_portion",
  "conventional_price_lb",
  "sustainable_price_lb",
  "default_sus",
  "conventional_ghg_per_lb",
  "baseline_freq"
)

required_cols <- c(
  "expected_lb_meat_portion",
  "conventional_price_lb",
  "sustainable_price_lb",
  "default_sus",
  "conventional_ghg_per_lb",
  "baseline_freq"
)

missing_cols <- setdiff(required_cols, names(opt))
if (length(missing_cols) > 0) {
  stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))
}

if (any(is.na(opt$expected_lb_meat_portion))) {
  stop("expected_lb_meat_portion has NA values in optimization_data_cafe3.")
}

if (any(opt$expected_lb_meat_portion < 0)) {
  stop("expected_lb_meat_portion has negative values in optimization_data_cafe3.")
}

if (any(is.na(opt$default_sus))) {
  stop("default_sus has NA values in optimization_data_cafe3.")
}

if (any(is.na(opt$conventional_ghg_per_lb))) {
  stop("conventional_ghg_per_lb has NA values in optimization_data_cafe3.")
}

default_sus_clean <- tolower(trimws(as.character(opt$default_sus)))
sus_flag <- as.integer(default_sus_clean == "yes")

if (any(!(default_sus_clean %in% c("yes", "no")))) {
  stop("default_sus contains values other than Yes/No.")
}

if (any(sus_flag == 1 & is.na(opt$sustainable_price_lb))) {
  stop("Some default-sustainable ingredients are missing sustainable_price_lb.")
}

if (any(sus_flag == 0 & is.na(opt$conventional_price_lb))) {
  stop("Some default-conventional ingredients are missing conventional_price_lb.")
}

default_sus_clean <- tolower(trimws(as.character(opt$default_sus)))
sus_flag <- as.integer(default_sus_clean == "yes")

# Effective price per lb used for costing
price_lb <- dplyr::case_when(
  sus_flag == 1 ~ opt$sustainable_price_lb,
  sus_flag == 0 ~ opt$conventional_price_lb
)

# Per-appearance quantities
portion_lb <- opt$expected_lb_meat_portion
cost_per_appearance <- price_lb * portion_lb
sus_cost_per_appearance <- dplyr::case_when(
  sus_flag == 1 ~ opt$sustainable_price_lb * portion_lb,
  sus_flag == 0 ~ 0,
  TRUE ~ NA_real_
)

ghg_per_appearance <- opt$conventional_ghg_per_lb * portion_lb

# -------------------------
# Constraint matrix
# -------------------------
A <- matrix(1, nrow = 1, ncol = n)
dir <- "="
rhs <- meals_served

A <- rbind(A, diag(n))
dir <- c(dir, rep(">=", n))
rhs <- c(rhs, lower)

if (any(!is.na(upper))) {
  idx <- which(!is.na(upper))
  A <- rbind(A, diag(n)[idx, , drop = FALSE])
  dir <- c(dir, rep("<=", length(idx)))
  rhs <- c(rhs, upper[idx])
}

# -------------------------
# Baseline totals
# -------------------------
baseline_cost <- sum(cost_per_appearance * x0)
baseline_sus  <- sum(sus_cost_per_appearance * x0)
baseline_ghg  <- sum(ghg_per_appearance * x0)

# -------------------------
# 5% cost reduction target
# -------------------------
target_cost <- baseline_cost * 0.95

A_costcap   <- rbind(A, cost_per_appearance)
dir_costcap <- c(dir, "<=")
rhs_costcap <- c(rhs, target_cost)


# -------------------------
# Safe numeric objects for lpSolve
# -------------------------

objective_stage1 <- as.numeric(sus_cost_per_appearance)

A_costcap_num <- matrix(
  as.numeric(A_costcap),
  nrow = nrow(A_costcap),
  ncol = ncol(A_costcap)
)

rhs_costcap_num <- as.numeric(rhs_costcap)
dir_costcap_chr <- as.character(dir_costcap)

if (any(is.na(objective_stage1) | is.nan(objective_stage1) | is.infinite(objective_stage1))) {
  stop("objective_stage1 contains NA/NaN/Inf.")
}

if (any(is.na(A_costcap_num) | is.nan(A_costcap_num) | is.infinite(A_costcap_num))) {
  stop("A_costcap_num contains NA/NaN/Inf.")
}

if (any(is.na(rhs_costcap_num) | is.nan(rhs_costcap_num) | is.infinite(rhs_costcap_num))) {
  stop("rhs_costcap_num contains NA/NaN/Inf.")
}

# =========================================================
# Stage 1: Maximize sustainable spend
# =========================================================
sol_s5_cafe3_stage1 <- lp(
  direction = "max",
  objective.in = objective_stage1,
  const.mat = A_costcap_num,
  const.dir = dir_costcap_chr,
  const.rhs = rhs_costcap_num,
  all.int = TRUE
)

if (sol_s5_cafe3_stage1$status != 0) {
  stop("s5_cafe3 Stage 1 infeasible. Try relaxing bounds or checking the 10% cost reduction target.")
}

x_s5_cafe3_sus <- sol_s5_cafe3_stage1$solution
opt_sus_s5_cafe3 <- sum(sus_cost_per_appearance * x_s5_cafe3_sus)

# =========================================================
# Stage 2: Minimize GHG while preserving max sustainability
# =========================================================
sus_slack <- 0.001

A_stage2   <- rbind(A_costcap, sus_cost_per_appearance)
dir_stage2 <- c(dir_costcap, ">=")
rhs_stage2 <- c(rhs_costcap, opt_sus_s5_cafe3 * (1 - sus_slack))

sol_s5_cafe3_stage2 <- lp(
  direction = "min",
  objective.in = ghg_per_appearance,
  const.mat = A_stage2,
  const.dir = dir_stage2,
  const.rhs = rhs_stage2,
  all.int = TRUE
)

if (sol_s5_cafe3_stage2$status != 0) {
  stop("s5_cafe3 Stage 2 infeasible. Try increasing sus_slack slightly.")
}

x_s5_cafe3_final <- sol_s5_cafe3_stage2$solution

# =========================================================
# Summaries
# =========================================================
summ <- function(x) {
  total_cost <- sum(cost_per_appearance * x)
  sus_spend  <- sum(sus_cost_per_appearance * x)
  sus_pct    <- if (total_cost > 0) sus_spend / total_cost else NA_real_
  total_ghg  <- sum(ghg_per_appearance * x)

  c(
    meals = sum(x),
    total_cost = total_cost,
    sus_spend = sus_spend,
    sus_pct = sus_pct,
    total_ghg = total_ghg
  )
}

s5_cafe3_table <- rbind(
  baseline = summ(x0),
  s5_cafe3_stage1_max_sus_given_10pct_cost_cut = summ(x_s5_cafe3_sus),
  s5_cafe3_final_min_ghg_given_10pct_cost_cut_and_max_sus = summ(x_s5_cafe3_final)
)

print(s5_cafe3_table)

# =========================================================
# Attach outputs back to optimization_data_cafe3
# =========================================================
optimization_data_cafe3 <- optimization_data_cafe3 %>%
  mutate(
    freq_baseline = baseline_freq,
    freq_opt_s5_cafe3 = x_s5_cafe3_final,
    delta_s5_cafe3 = freq_opt_s5_cafe3 - freq_baseline,

    baseline_cost_contribution = freq_baseline * cost_per_appearance,
    opt_cost_contribution_s5_cafe3 = freq_opt_s5_cafe3 * cost_per_appearance,

    baseline_sus_spend_contribution = freq_baseline * sus_cost_per_appearance,
    opt_sus_spend_contribution_s5_cafe3 = freq_opt_s5_cafe3 * sus_cost_per_appearance,

    baseline_ghg_contribution = freq_baseline * ghg_per_appearance,
    opt_ghg_contribution_s5_cafe3 = freq_opt_s5_cafe3 * ghg_per_appearance
  )

# -------------------------
# Helpful headline metrics
# -------------------------
opt_cost_s5_cafe3 <- sum(cost_per_appearance * x_s1_cafe3)
opt_sus_s5_cafe3  <- sum(sus_cost_per_appearance * x_s1_cafe3)
opt_ghg_s5_cafe3  <- sum(ghg_per_appearance * x_s1_cafe3)

pct_cost_reduction_s5_cafe3 <- if (baseline_cost > 0) {
  100 * (baseline_cost - opt_cost_s5_cafe3) / baseline_cost
} else {
  NA_real_
}

cat("\nBaseline cost: ", baseline_cost, "\n", sep = "")
cat("Optimized cost (s5_cafe3): ", opt_cost_s5_cafe3, "\n", sep = "")
cat("Percent cost reduction achieved: ", round(pct_cost_reduction_s5_cafe3, 2), "%\n", sep = "")
cat("Baseline sustainable spend: ", baseline_sus, "\n", sep = "")
cat("Optimized sustainable spend: ", opt_sus_s5_cafe3, "\n", sep = "")
cat("Baseline GHG: ", baseline_ghg, "\n", sep = "")
cat("Optimized GHG: ", opt_ghg_s5_cafe3, "\n", sep = "")

# =========================================================
# Category totals
# =========================================================
category_totals_s5_cafe3 <- optimization_data_cafe3 %>%
  group_by(category) %>%
  summarise(
    baseline_meals = sum(freq_baseline, na.rm = TRUE),
    opt_meals_s5_cafe3 = sum(freq_opt_s5_cafe3, na.rm = TRUE),
    delta_meals_s5_cafe3 = opt_meals_s5_cafe3 - baseline_meals,
    pct_change_meals_s5_cafe3 = ifelse(
      baseline_meals > 0,
      (delta_meals_s5_cafe3 / baseline_meals) * 100,
      NA_real_
    ),

    baseline_share = baseline_meals / sum(optimization_data_cafe3$freq_baseline, na.rm = TRUE),
    opt_share_s5_cafe3 = opt_meals_s5_cafe3 / sum(optimization_data_cafe3$freq_opt_s5_cafe3, na.rm = TRUE),

    baseline_cost = sum(baseline_cost_contribution, na.rm = TRUE),
    opt_cost_s5_cafe3 = sum(opt_cost_contribution_s5_cafe3, na.rm = TRUE),
    delta_cost_s5_cafe3 = opt_cost_s5_cafe3 - baseline_cost,
    pct_change_cost_s5_cafe3 = ifelse(
      baseline_cost > 0,
      (delta_cost_s5_cafe3 / baseline_cost) * 100,
      NA_real_
    ),

    baseline_sus_spend = sum(baseline_sus_spend_contribution, na.rm = TRUE),
    opt_sus_spend_s5_cafe3 = sum(opt_sus_spend_contribution_s5_cafe3, na.rm = TRUE),
    delta_sus_spend_s5_cafe3 = opt_sus_spend_s5_cafe3 - baseline_sus_spend,
    pct_change_sus_spend_s5_cafe3 = ifelse(
      baseline_sus_spend > 0,
      (delta_sus_spend_s5_cafe3 / baseline_sus_spend) * 100,
      NA_real_
    ),

    baseline_ghg = sum(baseline_ghg_contribution, na.rm = TRUE),
    opt_ghg_s5_cafe3 = sum(opt_ghg_contribution_s5_cafe3, na.rm = TRUE),
    delta_ghg_s5_cafe3 = opt_ghg_s5_cafe3 - baseline_ghg,
    pct_change_ghg_s5_cafe3 = ifelse(
      baseline_ghg > 0,
      (delta_ghg_s5_cafe3 / baseline_ghg) * 100,
      NA_real_
    ),

    baseline_sus_pct = ifelse(
      baseline_cost > 0,
      baseline_sus_spend / baseline_cost,
      NA_real_
    ),
    opt_sus_pct_s5_cafe3 = ifelse(
      opt_cost_s5_cafe3 > 0,
      opt_sus_spend_s5_cafe3 / opt_cost_s5_cafe3,
      NA_real_
    ),
    delta_sus_pct_s5_cafe3 = opt_sus_pct_s5_cafe3 - baseline_sus_pct,

    .groups = "drop"
  ) %>%
  arrange(desc(opt_meals_s5_cafe3))

category_totals_s5_cafe3

```


```{r}
# =========================================================
# s6_cafe3: Lock in 7% cost reduction, then maximize sustainability,
# then minimize GHG
# =========================================================

opt <- optimization_data_cafe3

x0 <- opt$baseline_freq
n  <- nrow(opt)

meals_served <- sum(x0)

# Lower and upper bounds
reduction <- 0.5
lower <- ceiling(x0 * reduction)

max_multiplier <- 1.5
upper <- floor(x0 * max_multiplier)

# -------------------------
# Feasibility checks
# -------------------------
required_cols <- c(
  "expected_lb_meat_portion",
  "conventional_price_lb",
  "sustainable_price_lb",
  "default_sus",
  "conventional_ghg_per_lb",
  "baseline_freq"
)

missing_cols <- setdiff(required_cols, names(opt))
if (length(missing_cols) > 0) {
  stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))
}

if (any(is.na(opt$expected_lb_meat_portion))) {
  stop("expected_lb_meat_portion has NA values in optimization_data_cafe3.")
}

if (any(opt$expected_lb_meat_portion < 0)) {
  stop("expected_lb_meat_portion has negative values in optimization_data_cafe3.")
}

if (any(is.na(opt$default_sus))) {
  stop("default_sus has NA values in optimization_data_cafe3.")
}

if (any(is.na(opt$conventional_ghg_per_lb))) {
  stop("conventional_ghg_per_lb has NA values in optimization_data_cafe3.")
}

default_sus_clean <- tolower(trimws(as.character(opt$default_sus)))

if (any(!(default_sus_clean %in% c("yes", "no")))) {
  stop("default_sus contains values other than Yes/No.")
}

sus_flag <- as.integer(default_sus_clean == "yes")

if (any(sus_flag == 1 & is.na(opt$sustainable_price_lb))) {
  stop("Some default-sustainable ingredients are missing sustainable_price_lb.")
}

if (any(sus_flag == 0 & is.na(opt$conventional_price_lb))) {
  stop("Some default-conventional ingredients are missing conventional_price_lb.")
}

# -------------------------
# Effective per-appearance values
# -------------------------
price_lb <- dplyr::case_when(
  sus_flag == 1 ~ opt$sustainable_price_lb,
  sus_flag == 0 ~ opt$conventional_price_lb
)

portion_lb <- opt$expected_lb_meat_portion

cost_per_appearance <- price_lb * portion_lb

sus_cost_per_appearance <- dplyr::case_when(
  sus_flag == 1 ~ opt$sustainable_price_lb * portion_lb,
  sus_flag == 0 ~ 0,
  TRUE ~ NA_real_
)

ghg_per_appearance <- opt$conventional_ghg_per_lb * portion_lb

# -------------------------
# Constraint matrix
# -------------------------
A <- matrix(1, nrow = 1, ncol = n)
dir <- "="
rhs <- meals_served

A <- rbind(A, diag(n))
dir <- c(dir, rep(">=", n))
rhs <- c(rhs, lower)

if (any(!is.na(upper))) {
  idx <- which(!is.na(upper))
  A <- rbind(A, diag(n)[idx, , drop = FALSE])
  dir <- c(dir, rep("<=", length(idx)))
  rhs <- c(rhs, upper[idx])
}

# -------------------------
# Baseline totals
# -------------------------
baseline_cost <- sum(cost_per_appearance * x0)
baseline_sus  <- sum(sus_cost_per_appearance * x0)
baseline_ghg  <- sum(ghg_per_appearance * x0)

# -------------------------
# 7% cost reduction target
# -------------------------
target_cost <- baseline_cost * 0.93

A_costcap   <- rbind(A, cost_per_appearance)
dir_costcap <- c(dir, "<=")
rhs_costcap <- c(rhs, target_cost)

# -------------------------
# Safe numeric objects for lpSolve
# -------------------------
objective_stage1 <- as.numeric(sus_cost_per_appearance)

A_costcap_num <- matrix(
  as.numeric(A_costcap),
  nrow = nrow(A_costcap),
  ncol = ncol(A_costcap)
)

rhs_costcap_num <- as.numeric(rhs_costcap)
dir_costcap_chr <- as.character(dir_costcap)

if (any(is.na(objective_stage1) | is.nan(objective_stage1) | is.infinite(objective_stage1))) {
  stop("objective_stage1 contains NA/NaN/Inf.")
}

if (any(is.na(A_costcap_num) | is.nan(A_costcap_num) | is.infinite(A_costcap_num))) {
  stop("A_costcap_num contains NA/NaN/Inf.")
}

if (any(is.na(rhs_costcap_num) | is.nan(rhs_costcap_num) | is.infinite(rhs_costcap_num))) {
  stop("rhs_costcap_num contains NA/NaN/Inf.")
}

# =========================================================
# Stage 1: Maximize sustainable spend given 7% cost cut
# =========================================================
sol_s6_cafe3_stage1 <- lp(
  direction = "max",
  objective.in = objective_stage1,
  const.mat = A_costcap_num,
  const.dir = dir_costcap_chr,
  const.rhs = rhs_costcap_num,
  all.int = TRUE
)

if (sol_s6_cafe3_stage1$status != 0) {
  stop("s6_cafe3 Stage 1 infeasible. Try relaxing bounds or checking the 7% cost reduction target.")
}

x_s6_cafe3_sus <- sol_s6_cafe3_stage1$solution
opt_sus_s6_cafe3 <- sum(sus_cost_per_appearance * x_s6_cafe3_sus)

# =========================================================
# Stage 2: Minimize GHG while preserving max sustainability
# =========================================================
sus_slack <- 0.001

A_stage2   <- rbind(A_costcap, sus_cost_per_appearance)
dir_stage2 <- c(dir_costcap, ">=")
rhs_stage2 <- c(rhs_costcap, opt_sus_s6_cafe3 * (1 - sus_slack))

sol_s6_cafe3_stage2 <- lp(
  direction = "min",
  objective.in = ghg_per_appearance,
  const.mat = A_stage2,
  const.dir = dir_stage2,
  const.rhs = rhs_stage2,
  all.int = TRUE
)

if (sol_s6_cafe3_stage2$status != 0) {
  stop("s6_cafe3 Stage 2 infeasible. Try increasing sus_slack slightly.")
}

x_s6_cafe3_final <- sol_s6_cafe3_stage2$solution

# =========================================================
# Summaries
# =========================================================
summ <- function(x) {
  total_cost <- sum(cost_per_appearance * x)
  sus_spend  <- sum(sus_cost_per_appearance * x)
  sus_pct    <- if (total_cost > 0) sus_spend / total_cost else NA_real_
  total_ghg  <- sum(ghg_per_appearance * x)

  c(
    meals = sum(x),
    total_cost = total_cost,
    sus_spend = sus_spend,
    sus_pct = sus_pct,
    total_ghg = total_ghg
  )
}

s6_cafe3_table <- rbind(
  baseline = summ(x0),
  s6_cafe3_stage1_max_sus_given_7pct_cost_cut = summ(x_s6_cafe3_sus),
  s6_cafe3_final_min_ghg_given_7pct_cost_cut_and_max_sus = summ(x_s6_cafe3_final)
)

print(s6_cafe3_table)

# =========================================================
# Attach outputs back to optimization_data_cafe3
# =========================================================
optimization_data_cafe3 <- optimization_data_cafe3 %>%
  mutate(
    freq_baseline = baseline_freq,
    freq_opt_s6_cafe3 = x_s6_cafe3_final,
    delta_s6_cafe3 = freq_opt_s6_cafe3 - freq_baseline,

    baseline_cost_contribution = freq_baseline * cost_per_appearance,
    opt_cost_contribution_s6_cafe3 = freq_opt_s6_cafe3 * cost_per_appearance,

    baseline_sus_spend_contribution = freq_baseline * sus_cost_per_appearance,
    opt_sus_spend_contribution_s6_cafe3 = freq_opt_s6_cafe3 * sus_cost_per_appearance,

    baseline_ghg_contribution = freq_baseline * ghg_per_appearance,
    opt_ghg_contribution_s6_cafe3 = freq_opt_s6_cafe3 * ghg_per_appearance
  )

# -------------------------
# Helpful headline metrics
# -------------------------
opt_cost_s6_cafe3 <- sum(cost_per_appearance * x_s6_cafe3_final)
opt_sus_s6_cafe3  <- sum(sus_cost_per_appearance * x_s6_cafe3_final)
opt_ghg_s6_cafe3  <- sum(ghg_per_appearance * x_s6_cafe3_final)

pct_cost_reduction_s6_cafe3 <- if (baseline_cost > 0) {
  100 * (baseline_cost - opt_cost_s6_cafe3) / baseline_cost
} else {
  NA_real_
}

pct_sus_change_s6_cafe3 <- if (baseline_cost > 0) {
  100 * (baseline_sus - opt_sus_s6_cafe3) / baseline_sus
} else {
  NA_real_
}

cat("\nBaseline cost: ", baseline_cost, "\n", sep = "")
cat("Optimized cost (s6_cafe3): ", opt_cost_s6_cafe3, "\n", sep = "")
cat("Percent cost reduction achieved: ", round(pct_cost_reduction_s6_cafe3, 2), "%\n", sep = "")
cat("Baseline sustainable spend: ", baseline_sus, "\n", sep = "")
cat("Optimized sustainable spend: ", opt_sus_s6_cafe3, "\n", sep = "")
cat("Percent sustainable spend change: ", pct_sus_change_s6_cafe3, "\n", sep = "")
cat("Baseline GHG: ", baseline_ghg, "\n", sep = "")
cat("Optimized GHG: ", opt_ghg_s6_cafe3, "\n", sep = "")

# =========================================================
# Category totals
# =========================================================
category_totals_s6_cafe3 <- optimization_data_cafe3 %>%
  group_by(category) %>%
  summarise(
    baseline_meals = sum(freq_baseline, na.rm = TRUE),
    opt_meals_s6_cafe3 = sum(freq_opt_s6_cafe3, na.rm = TRUE),
    delta_meals_s6_cafe3 = opt_meals_s6_cafe3 - baseline_meals,
    pct_change_meals_s6_cafe3 = ifelse(
      baseline_meals > 0,
      (delta_meals_s6_cafe3 / baseline_meals) * 100,
      NA_real_
    ),

    baseline_share = baseline_meals / sum(optimization_data_cafe3$freq_baseline, na.rm = TRUE),
    opt_share_s6_cafe3 = opt_meals_s6_cafe3 / sum(optimization_data_cafe3$freq_opt_s6_cafe3, na.rm = TRUE),

    baseline_cost = sum(baseline_cost_contribution, na.rm = TRUE),
    opt_cost_s6_cafe3 = sum(opt_cost_contribution_s6_cafe3, na.rm = TRUE),
    delta_cost_s6_cafe3 = opt_cost_s6_cafe3 - baseline_cost,
    pct_change_cost_s6_cafe3 = ifelse(
      baseline_cost > 0,
      (delta_cost_s6_cafe3 / baseline_cost) * 100,
      NA_real_
    ),

    baseline_sus_spend = sum(baseline_sus_spend_contribution, na.rm = TRUE),
    opt_sus_spend_s6_cafe3 = sum(opt_sus_spend_contribution_s6_cafe3, na.rm = TRUE),
    delta_sus_spend_s6_cafe3 = opt_sus_spend_s6_cafe3 - baseline_sus_spend,
    pct_change_sus_spend_s6_cafe3 = ifelse(
      baseline_sus_spend > 0,
      (delta_sus_spend_s6_cafe3 / baseline_sus_spend) * 100,
      NA_real_
    ),

    baseline_ghg = sum(baseline_ghg_contribution, na.rm = TRUE),
    opt_ghg_s6_cafe3 = sum(opt_ghg_contribution_s6_cafe3, na.rm = TRUE),
    delta_ghg_s6_cafe3 = opt_ghg_s6_cafe3 - baseline_ghg,
    pct_change_ghg_s6_cafe3 = ifelse(
      baseline_ghg > 0,
      (delta_ghg_s6_cafe3 / baseline_ghg) * 100,
      NA_real_
    ),

    baseline_sus_pct = ifelse(
      baseline_cost > 0,
      baseline_sus_spend / baseline_cost,
      NA_real_
    ),
    opt_sus_pct_s6_cafe3 = ifelse(
      opt_cost_s6_cafe3 > 0,
      opt_sus_spend_s6_cafe3 / opt_cost_s6_cafe3,
      NA_real_
    ),
    delta_sus_pct_s6_cafe3 = opt_sus_pct_s6_cafe3 - baseline_sus_pct,

    .groups = "drop"
  ) %>%
  arrange(desc(opt_meals_s6_cafe3))

category_totals_s6_cafe3

## Save data
write.csv(optimization_data_cafe3, "Optimized_freqs.csv", row.names = FALSE)


```



## BELOW IS GENERIC not C-3 specific

Data load

```{r}
safe_mean <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

# clean column names using janitor
meals <- read.csv("../Basic_Data/F25_Sp26_meals.csv", stringsAsFactors = FALSE) %>%
  clean_names()

recipe_ingredient_dict <- read.csv("../Basic_Data/Recipe-Ingredient_dictionary.csv", stringsAsFactors = FALSE) %>%
  clean_names()

ingredient_prices <- read.csv("../Basic_Data/Ingredient_prices.csv", stringsAsFactors = FALSE) %>%
  clean_names()

ghg_equivalents <- read.csv("../Basic_Data/GHG_equivalents.csv", stringsAsFactors = FALSE) %>%
  clean_names()
```

Build this into a sheet called optimization_data which we can use for the optimization

```{r}

# -------------------------
# Clean / standardize values
# -------------------------
meals <- meals %>%
  mutate(
    ingredient = trimws(ingredient),
    category = trimws(category),
    planned_portions = as.numeric(gsub(",", "", planned_portions)),
    percent_meat = as.numeric(percent_meat),
    meat_price_per_dish = as.numeric(meat_price_per_dish),
    expected_lb_meat_portion = as.numeric(expected_lb_meat_portion),
    oz_meat_per_dish = as.numeric(oz_meat_per_dish),
    portion_size_oz = as.numeric(portion_size_oz),
    planned_weight_lbs = as.numeric(planned_weight_lbs),
    cost_recipe_per_portion = as.numeric(cost_recipe_per_portion)
  )

ingredient_prices <- ingredient_prices %>%
  mutate(
    ingredient = trimws(ingredient),
    category = trimws(category),
    conventional_price_lb = as.numeric(conventional_price_lb),
    sustainable_price_lb = as.numeric(sustainable_price_lb),
    default_sus = trimws(default_sus)
  )

ghg_equivalents <- ghg_equivalents %>%
  mutate(
    meat_type = trimws(meat_type),
    c_footprint_kg_c_per_kg_food = as.numeric(c_footprint_kg_c_per_kg_food)
  )

# -------------------------
# Summarise recipe-level variables to ingredient level
# -------------------------
ingredient_summary <- meals %>%
  filter(!is.na(ingredient), ingredient != "") %>%
  group_by(ingredient) %>%
  summarise(
    baseline_freq = n(),
    percent_dish_meat = safe_mean(percent_meat),
    meat_price_per_dish = safe_mean(meat_price_per_dish),
    expected_lb_meat_portion = safe_mean(expected_lb_meat_portion),
    oz_meat_per_dish = safe_mean(oz_meat_per_dish),
    portion_size_oz = safe_mean(portion_size_oz),
    expected_portions = safe_mean(planned_portions),
    planned_weight_lbs = safe_mean(planned_weight_lbs),
    cost_recipe_per_portion = safe_mean(cost_recipe_per_portion),
    category_meals = dplyr::first(na.omit(category)),
    .groups = "drop"
  )

# -------------------------
# Ingredient-level lookup fields
# -------------------------
ingredient_lookup <- ingredient_prices %>%
  distinct(ingredient, .keep_all = TRUE) %>%
  transmute(
    ingredient,
    category_prices = category,
    conventional_price_lb,
    sustainable_price_lb,
    default_sus
  )

# -------------------------
# GHG lookup
# -------------------------
ghg_lookup <- ghg_equivalents %>%
  distinct(meat_type, .keep_all = TRUE) %>%
  transmute(
    category = meat_type,
    conventional_ghg_per_lb = c_footprint_kg_c_per_kg_food
  )

# -------------------------
# Final optimization dataset
# -------------------------
optimization_data <- ingredient_summary %>%
  left_join(ingredient_lookup, by = "ingredient") %>%
  mutate(
    category = coalesce(category_prices, category_meals)
  ) %>%
  left_join(ghg_lookup, by = "category") %>%
  select(
    ingredient,
    category,
    percent_dish_meat,
    meat_price_per_dish,
    expected_lb_meat_portion,
    oz_meat_per_dish,
    conventional_price_lb,
    sustainable_price_lb,
    default_sus,
    portion_size_oz,
    expected_portions,
    planned_weight_lbs,
    cost_recipe_per_portion,
    conventional_ghg_per_lb,
    baseline_freq
  )

# -------------------------
# Diagnostic checks
# -------------------------

# 1. Missing price lookups
missing_prices <- optimization_data %>%
  filter(is.na(conventional_price_lb) | is.na(sustainable_price_lb))

# 2. Missing GHG lookups
missing_ghg <- optimization_data %>%
  filter(is.na(conventional_ghg_per_lb))

# 3. Ingredients with multiple meal-side categories
category_conflicts <- meals %>%
  filter(!is.na(ingredient), ingredient != "") %>%
  group_by(ingredient) %>%
  summarise(
    n_categories = n_distinct(category[!is.na(category) & category != ""]),
    .groups = "drop"
  ) %>%
  filter(n_categories > 1)

# 4. Compare meal-side vs price-side category
category_compare <- ingredient_summary %>%
  left_join(ingredient_lookup, by = "ingredient") %>%
  filter(
    !is.na(category_meals),
    !is.na(category_prices),
    category_meals != category_prices
  ) %>%
  select(ingredient, category_meals, category_prices)

# View diagnostics
missing_prices
missing_ghg
category_conflicts
category_compare

```

